---
title: "Activity02_SLR"
author: "Maureen Muthengi"
date: "`r Sys.Date()`"
output:
  word_document: default
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Loading Libraries
```{r}
library(tidyverse)
library(tidymodels)
library(skimr)
library(ggplot2)
library(broom)
library(dbplyr)
library(e1071)  # For skewness calculation
```

#Load Data
```{r}
hfi <-read_csv("hfi.csv")
hfi
```
## 1.  What are the dimensions of the dataset? 
## There are 1458 rows and 123 columns

## What does each row represent? The observations for example year, ISO_code, countries etc.

```{r}
glimpse(hfi)
```

## The dataset spans a lot of years. We are only interested in data from year 2016. In the R code chunk below titled `hfi-2016`, type the code that does the following:

## -   Filter the data `hfi` data frame for year 2016, and
##v-   Assigns the result to a data frame named `hfi_2016`.
```{r}
hfi_2016 <- hfi %>%
  filter(year == 2016)
hfi_2016

```
### 1. Identify our research question(s)

## The research question is often defined by you (or your company, boss, etc.). Today's research question/goal is to predict a country's personal freedom score in 2016.

## For this activity we want to explore the relationship between the personal freedom score, `pf_score`, and the political pressures and controls on media content index,`pf_expression_control`. Specifically, we are going to use the political pressures and controls on media content index to predict a country's personal freedom score in 2016.

### 2. Explore the variables of interest
#Answer the following questions (use your markdown skills) and complete the following tasks.

#2.  What type of plot would you use to display the distribution of the personal freedom scores, `pf_score`? Would this be the same type of plot to display the distribution of the political pressures and controls on media content index, `pf_expression_control`?

-   In the R code chunk below titled `univariable-plots`, type the R code that displays this plot for `pf_score`.
-   In the R code chunk below titled `univariable-plots`, type the R code that displays this plot for `pf_expression_control`.

```{r}
# Plot the distribution of pf_score
ggplot(hfi_2016, aes(x = pf_score)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Personal Freedom Scores", x = "Personal Freedom Score", y = "Frequency") +
  theme_minimal()

# Plot the distribution of pf_expression_control
ggplot(hfi_2016, aes(x = pf_expression_control)) +
  geom_histogram(binwidth = 1, fill = "red", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Political Pressures and Controls on Media Content", x = "Political Pressures and Controls on Media Content Index", y = "Frequency") +
  theme_minimal()
```
## Comment on each of these two distributions. Be sure to describe their centers, spread, shape, and any potential outliers.
## the distribution of pf scores is left skewed and the center is at 5 while the the distribution of pf_expression_control is normally distributed, the center ist at 4 and there are no outliers.

## What type of plot would you use to display the relationship between the personal freedom score, `pf_score`, and the political pressures and controls on media content index,`pf_expression_control`? A scatter plot
## There is a linear relationship between the personal freedom score, `pf_score`, and the political pressures and controls on media content index,`pf_expression_control`

```{r}
# `relationship-plot`
ggplot(hfi_2016, aes(x = pf_expression_control, y = pf_score)) +
  geom_point(color = "blue", alpha = 0.6) +
  labs(title = "Relationship between Personal Freedom Score and Political Pressures and Controls on Media Content",
       x = "Political Pressures and Controls on Media Content Index",
       y = "Personal Freedom Score") +
  theme_minimal()
```
## 4.  Does the relationship look linear? Yes it looks linear
## If you knew a country's `pf_expression_control`, or its score out of 10, with 0 being the most, of political pressures and controls on media content, would you be comfortable using a linear model to predict the personal freedom score? Yes

#### Challenge

For each plot and using your `{dplyr}` skills, obtain the appropriate numerical summary statistics and provide more detailed descriptions of these plots. For example, in (4) you were asked to comment on the center, spread, shape, and potential outliers. What measures could/should be used to describe these? You might not know of one for each of those terms.

What numerical summary would you use to describe the relationship between two numerical variables? (hint: explore the `cor` function from Base R)

```{r challenge}
# Summary statistics for pf_score
pf_score_summary <- hfi_2016 %>%
  summarise(
    mean = mean(pf_score, na.rm = TRUE),
    median = median(pf_score, na.rm = TRUE),
    sd = sd(pf_score, na.rm = TRUE),
    IQR = IQR(pf_score, na.rm = TRUE),
    skewness = skewness(pf_score, na.rm = TRUE),
    min = min(pf_score, na.rm = TRUE),
    max = max(pf_score, na.rm = TRUE)
  )

# Identifying outliers for pf_score using 1.5 * IQR rule
Q1_pf_score <- quantile(hfi_2016$pf_score, 0.25, na.rm = TRUE)
Q3_pf_score <- quantile(hfi_2016$pf_score, 0.75, na.rm = TRUE)
IQR_pf_score <- Q3_pf_score - Q1_pf_score
outliers_pf_score <- hfi %>%
  filter(pf_score < (Q1_pf_score - 1.5 * IQR_pf_score) | pf_score > (Q3_pf_score + 1.5 * IQR_pf_score))

# Summary statistics for pf_expression_control
pf_expression_control_summary <- hfi_2016 %>%
  summarise(
    mean = mean(pf_expression_control, na.rm = TRUE),
    median = median(pf_expression_control, na.rm = TRUE),
    sd = sd(pf_expression_control, na.rm = TRUE),
    IQR = IQR(pf_expression_control, na.rm = TRUE),
    skewness = skewness(pf_expression_control, na.rm = TRUE),
    min = min(pf_expression_control, na.rm = TRUE),
    max = max(pf_expression_control, na.rm = TRUE)
  )

# Identifying outliers for pf_expression_control using 1.5 * IQR rule
Q1_pf_expression_control <- quantile(hfi_2016$pf_expression_control, 0.25, na.rm = TRUE)
Q3_pf_expression_control <- quantile(hfi_2016$pf_expression_control, 0.75, na.rm = TRUE)
IQR_pf_expression_control <- Q3_pf_expression_control - Q1_pf_expression_control
outliers_pf_expression_control <- hfi %>%
  filter(pf_expression_control < (Q1_pf_expression_control - 1.5 * IQR_pf_expression_control) | pf_expression_control > (Q3_pf_expression_control + 1.5 * IQR_pf_expression_control))

# Calculate correlation between pf_score and pf_expression_control
correlation <- cor(hfi_2016$pf_score, hfi_2016$pf_expression_control, use = "complete.obs")
correlation

# Print results
print(pf_score_summary)
print(outliers_pf_score)
print(pf_expression_control_summary)
print(outliers_pf_expression_control)
print(correlation)
```
```{r}
print(correlation)
```
## The strong positive correlation coefficient of 0.8450646 indicates that personal freedom scores and political pressures and controls on media content index are strongly positively related. This suggests that in countries where personal freedoms are higher, political pressures and controls on media content are generally lower, demonstrating a significant linear relationship between these two variables. This relationship is visually supported by the scatter plot showing a clear upward trend.

## Implications: The high correlation suggests that countries with higher personal freedom scores tend to have lower levels of political pressure and control on media content. This could imply that in environments where personal freedoms are respected, there is less political interference in media content.

### 3. Fit a simple linear regression model

## Regardless of your response to (4), we will continue fitting a simple linear regression (SLR) model to these data. The code that we will be using to fit statistical models in this course use `{tidymodels}` - an opinionated way to fit models in R - and this is likely new to most of you. I will provide you with example code when I do not think you should know what to do - i.e., anything `{tidymodels}` related.

## To begin, we will create a `{parsnip}` specification for a linear model.

```{r parsnip-spec}
lm_spec <- linear_reg() %>%
  set_mode("regression") %>%
  set_engine("lm")

lm_spec
```
## Using this specification, we can now fit our model: $\texttt{pf\\_score} = \beta_0 + \beta_1 \times \texttt{pf\\_expression\\_control} + \varepsilon$. Note, the "\$" portion in the previous sentence is LaTeX snytex which is a math scripting (and other scripting) language. I do not expect you to know this, but you will become more comfortable with this. Look at your knitted document to see how this syntax appears.

## In the code chunk below titled `fit-lm`, replace "verbatim" with "r" just before the code chunk title.

```{r fit-lm}
slr_mod <- lm_spec %>% 
  fit(pf_score ~ pf_expression_control, data = hfi_2016)

tidy(slr_mod)
```
## Intercept: 4.2838153 This is the estimated personal freedom score when the political pressures and controls on media content index is 0. It represents the baseline level of personal freedom when there are no political pressures or controls on media content.
## Standard Error: 0.14897525 This indicates the variability in the estimate of the intercept. A smaller standard error suggests more precise estimation.
## t-Statistic: 28.75521, the t-statistic for the intercept tests the null hypothesis that the intercept is zero. A large absolute t-value indicates that the intercept is significantly different from zero.
## p-Value: 4.228325×10−65, the p-value is extremely small, suggesting that the intercept is highly statistically significant. We can reject the null hypothesis that the intercept is zero.
## Slope- pf_expression_control): 0.5418452, this is the estimated change in the personal freedom score for each one-unit increase in the political pressures and controls on media content index. A positive estimate indicates that as political pressures and controls increase, the personal freedom score also increases.
## Standard Error: 0.02710231, this measures the precision of the slope estimate. A smaller standard error indicates a more precise estimate.
## t-Statistic: 19.99258, the t-statistic for the slope tests the null hypothesis that the slope is zero. A large absolute t-value suggests that the slope is significantly different from zero.
## p-Value: 2.313214×10−45, the p-value is extremely small, indicating that the slope is highly statistically significant. We can reject the null hypothesis that the slope is zero.
## Overall Model Interpretation: The intercept and the slope are both highly statistically significant, with p-values much smaller than the common significance level of 0.05.
## The positive slope 0.5418452 indicates a strong positive linear relationship between pf_expression_control and pf_score. Specifically, for each one-unit increase in the political pressures and controls on media content index, the personal freedom score increases by approximately 0.542 units. Given the high t-statistics and very low p-values, we can conclude that both the intercept and the slope are significant contributors to the model, providing strong evidence of a relationship between the political pressures and controls on media content index and the personal freedom score.
## Conclusion: The model indicates that as political pressures and controls on media content increase, personal freedom scores also increase. This might initially seem counterintuitive, but it could suggest that in contexts where there are more political pressures on media, there might be an increased effort or mechanism to ensure personal freedoms are maintained or improved. Further investigation would be necessary to understand the underlying causes of this relationship.

## Using the tidy output, the estimated parameters are:

## Intercept: 4.2838153
## Slope: 0.5418452
## Therefore, the updated formula is:

## pf_score^=4.2838153 + 0.5418452 × pf_expression_control
## pf_score = 4.2838153 + 0.5418452 × pf_expression_control this formula can now be used to predict the personal freedom score based on the value of the political pressures and controls on media content index.

## Intercept 4.2838153, the intercept represents the estimated personal freedom score when the political pressures and controls on media content index (\texttt{pf_expression_control}) is 0. If there were no political pressures and controls on media content (i.e., \texttt{pf_expression_control} = 0), the personal freedom score would be approximately 4.28. This provides a baseline level of personal freedom in the absence of political pressures on media content.
## Slope 0.5418452, the slope represents the estimated change in the personal freedom score for each one-unit increase in the political pressures and controls on media content index (\texttt{pf_expression_control}). Specifically, for every one-unit increase in the \texttt{pf_expression_control} index, the personal freedom score is expected to increase by approximately 0.542 units.
## This positive relationship suggests that higher political pressures and controls on media content are associated with higher personal freedom scores. This might seem counterintuitive at first, but it could imply that in environments where there are more political pressures and controls on media, there could be compensatory mechanisms or efforts in place to ensure personal freedoms are maintained or improved.
## The intercept provides a starting point for personal freedom scores in the theoretical scenario where political pressures and controls on media content are completely absent.
## The slope indicates the direction and strength of the relationship between political pressures on media and personal freedom scores. The positive slope suggests a direct, although not necessarily causal, relationship where increasing political pressures and controls on media content correlate with increasing personal freedom scores.

## Summary > For countries with a `pf_expression_control` of 0 (those with the largest amount of political pressure on media content), we expect their mean personal freedom score to be 4.28. > For every 1 unit increase in `pf_expression_control` (political pressure on media content index), we expect a country's mean personal freedom score to increase 0.542 units.

### 4. Assessing

#### 4.A: Assess with your Day 1 model

## To assess our model fit, we can use $R^2$ (the coefficient of determination), the proportion of variability in the response variable that is explained by the explanatory variable. We use `glance` from `{broom}` (which is automatically loaded with `{tidymodels}` - `{broom}` is also where `tidy` is from) to access this information.

```{r glance-mod}
glance(slr_mod)
```
## 7.  What is the value of $R^2$ for this model? 0.7141342 or 71.41%

## 8.  What does this value mean in the context of this model? Think about what would a "good" value of $R^2$ would be? Can/should this value be "perfect"? This means 71.41% of the variation in personal freedom scores across different countries can be accounted for by the variation in political pressures and controls on media content. This suggests that the model captures a significant portion that is 71.41% of the relationship between political pressures on media content and personal freedom scores.

#### 4.B: Assess with test/train

##You previously fit a model and evaluated it using the exact same data. This is a bit of circular reasoning and does not provide much information about the model's performance. Now we will work through the test/train process of fitting and assessing a simple linear regression model.


## -   Create a new R code chunk and provide it with a descriptive tile (e.g., `train-test`).
## -   Set a seed.
## -   Create an initial 80-20 split of the `hfi_2016` dataset
## -   Using your initial split R object, assign the two splits into a training R object and a testing R object.

## Now, you will use your training dataset to fit a SLR model.
```{r}
#train-test
# set seed before random split
set.seed(1)
# put 80% of the data into the training set
slr_split <- initial_split(hfi_2016, prop = 0.80)

# assign the two splits to data frames - with descriptive names
slr_train <- training(slr_split)
slr_test <- testing(slr_split)

# splits
slr_train

```


```{r train-fit-lm}
slr_train <- lm_spec %>% 
  fit(pf_score ~ pf_expression_control, data = hfi_2016)

tidy(slr_train)
```

## Notice that you can reuse the `lm_spec` specification because we are still doing a linear model.

## 9.  Using the `tidy` output, update the below formula with the estimated parameters. That is, replace "intercept" and "slope" with the appropriate values

$\widehat{\texttt{pf\\_score}} = intercept + slope \times \texttt{pf\\_expression\\_control}$

## 10. Interpret each of the estimated parameters from (10) in the context of this research question. That is, what do these values represent?

## Now we will assess using the testing data set.

```{r glance-test}
#train-fit-lm
test_aug <- augment(slr_train, new_data = slr_test)
test_aug
```

## Look at the various information produced by this code. Can you identify what each column represents?

## The `.pred` column in this output can also be obtained by using `predict` (i.e., `predict(slr_fit, new_data = data_test)`)
## .pred: This column contains the predicted values generated by the linear model for the response variable.

## .resid: This column contains the residuals, which are the differences between the actual observed values and the predicted values (.resid = observed - .pred).

## year: This column contains the year associated with each observation.

## ISO_code: This column contains the ISO country codes, which are typically three-letter country abbreviations.

## countries: This column contains the full names of the countries corresponding to each observation.

## region: This column contains the region classifications for each country (e.g., Latin America & the Caribbean, Caucasus & Central Asia, etc.).

## pf_rol_procedural: This column contains the observed values of the response variable being predicted by the linear model. It represents the procedural component of the rule of law index for each country and year.

## pf_rol_civil: This column contains values representing the procedural component of the rule of law index, specifically related to civil justice.

## pf_rol_criminal: This column contains values representing the procedural component of the rule of law index, specifically related to criminal justice.

## pf_rol: This column contains the overall procedural component of the rule of law index, which might be an aggregate or composite score derived from various subcomponents, including civil and criminal justice.

## pf_ss_homicide: This column contains values representing the homicide rate or an index related to homicides, which is a component of safety and security.

## pf_ss_disappearances_disap: This column contains values representing the rate or index related to enforced disappearances, indicating incidents where individuals have been disappeared by state or non-state actors.

## pf_ss_disappearances_violent: This column contains values representing the rate or index related to violent disappearances, which may include kidnappings or disappearances resulting from violence.

### Model diagnostics

### To assess whether the linear model is reliable, we should check for (1) linearity, (2) nearly normal residuals, and (3) constant variability. Note that the normal residuals is not really necessary for all models (sometimes we simply want to describe a relationship for the data that we have or population-level data, where statistical inference is not appropriate/necessary).

### In order to do these checks we need access to the fitted (predicted) values and the residuals. We can use `broom::augment` to calculate these.

```{r augment}
train_aug <- augment(slr_mod, new_data = hfi_2016)
print(train_aug)
```
## **Linearity**: You already checked if the relationship between `pf_score` and `pf_expression_control` is linear using a scatterplot. We should also verify this condition with a plot of the residuals vs. fitted (predicted) values.

```{r fitted-residual}
ggplot(data = train_aug, aes(x = .pred, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  xlab("Fitted values") +
  ylab("Residuals")
```
### Notice here that `train_aug` can also serve as a data set because stored within it are the fitted values ($\hat{y}$) and the residuals. Also note that we are getting fancy with the code here. After creating the scatterplot on the first layer (first line of code), we overlay a red horizontal dashed line at $y = 0$ (to help us check whether the residuals are distributed around 0), and we also rename the axis labels to be more informative.

### Answer the following question:

### 11. Is there any apparent pattern in the residuals plot? Pattern: Residuals are randomly scattered around the horizontal line (y = 0) and don't form any specific pattern. What does this indicate about the linearity of the relationship between the two variables? The relationship between the predictors and the response variable are not linear.

### **Nearly normal residuals**: To check this condition, we can look at a histogram of the residuals.

```{r}
ggplot(data = train_aug, aes(x = .resid)) +
  geom_histogram(binwidth = 0.25) +
  xlab("Residuals")
```
### Answer the following question:

### 12. Based on the histogram, does the nearly normal residuals condition appear to be violated? the residuals are approximately normally distributed, supporting the normality assumption.

### **Constant variability**:

### 13. Based on the residuals vs. fitted plot, does the constant variability condition appear to be violated? The spread (variability) of the residuals is roughly constant across all levels of fitted values. The residuals fan out as fitted values increase, the variability of the residuals is not constant.

## Attribution

## This document is based on labs from [OpenIntro](https://www.openintro.org/).